Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-03 Manuscript/manuscript.Rmd
manuscript edits; reordering figures

3. **Log2 Distance**: For each genomic bin, the $log_{2}$ of the distance in bases (refered to hereafter simply as $\textit{distance}$), from the center of the bin to the center of the nearest region defined by a specific genomic annotation was calculated. The log transformation was used as a normalization technique in order to account for the skewness present in the distance densities (Supplementary Figures 1-2).

@@

To systematically summarize and compare our results of data driven random forest models, we created an ensemble framework (Figure 1). Each system in the framework was assumed to be composed of a model constructed from a particular cell line, $L = \{GM12878, K562\}$. Within each cell line we evaluated 4 data resolutions, $R = \{10 kb, 25 kb, 50 kb, 100 kb\}$. We then considered 3 separate predictor types as the feature space including $P = \{OC, OP, Distance\}$. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including $B = \{None, RUS, ROS, SMOTE\}$. Each combination of the above parameters was implemented on chromosome specific sections of the linear genome. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9). Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.

@@

## IMBALANCE RATIOS REMAINED SEVERE irrespectively of RESOLUTION of Hi-C data

We analyzed publically available Hi-C data for the GM12878 and K562 cell lines provided by Rao et al. [@rao20143d]. Topologically associating domains were then identified using the Arrowhead algorithm [@durand2016juicebox] obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolutions, for each cell line respectively. We then established a unified approach toward preparing cell line specific TAD boundary data for predictive modelling purposes, irrespective of data resolutions (Figure 2).

@@

## HIGH JACCARD INDICES IMPLY COLOCALIZATION OF MANY GROUP-SPECIFC ANNOTATIONS

@@

Using the Jaccard index, we found that specific sets of these cell line-specific annotations were highly colocalized together on the linear genome (Table 3). For GM12878, the range of Jaccard indices for the 10 histone modifications was found to be (0.022-0.539). Specifically, H3k27ac, H3k4me1, H3k4me2, and H3k4me3 were found to be highly correlated, high levels of which have been detected surrounding transcription start sites [@barski2007high]. Likewise, for the 79 TFBS, the range of Jaccard indices was (<0.001-0.425). Here, known mediators of loop extrusion components were found to have high pairwise correlations including CTCF, RAD21, and SMC3 [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Similar profiles of Jaccard indices were exhibited across the sets of annotations for the K562 cell line (Supplementary Table 4). [??? So what? Why did you talk about it?; !!!added implication]. The high Jaccard indices among select sets of genomic annotations indicated the colocalization of annotations known to be enriched at TAD boundaries. This gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space of all annotations while retaining the possibly correlated annotations that could be predictive of TAD boundary formation.

To establish said feature space for all downstream random forest models, we constructed 3 different predictor-types. These predictor-type specific annotations represented the data matrix, $X = \{X_1, X_2, ..., X_p\}$, where $p$ was the complete set of functional annotations. Overlap counts was the least granular predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 3A). Overlap percents was a more refined predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 3B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 3C).

>>>>

3. **Log2 Distance**: For each genomic bin, the $log_{2}$ of the distance in bases (refered to hereafter simply as $\textit{distance}$), from the center of the bin to the center of the nearest region defined by a specific genomic annotation was calculated.

@@

@@

## IMBALANCE RATIOS REMAINED SEVERE IRRESPECTIVELY OF Hi-C DATA RESOLUTION

We analyzed publically available Hi-C data for the GM12878 and K562 cell lines provided by Rao et al. [@rao20143d]. Topologically associating domains were identified using the Arrowhead algorithm [@durand2016juicebox] obtained at 10 kb, 25 kb, 50 kb, and 100 kb resolutions, for each cell line respectively. We then established a unified approach toward preparing cell line specific TAD boundary data for predictive modelling purposes, irrespective of data resolutions (Figure 1).

@@

<!--
## HIGH JACCARD INDICES IMPLY COLOCALIZATION OF MANY GROUP-SPECIFC ANNOTATIONS 
-->

## ESTABLISHING THE FEATURE SPACE USING INCREASINGLY GRANULAR RELATIONSHIPS BETWEEN FUNCTIONAL ANNOTATIONS AND GENOMIC BINS

@@

<!--
Using the Jaccard index, we found that specific sets of these cell line-specific annotations were highly colocalized together on the linear genome (Table 3). For GM12878, the range of Jaccard indices for the 10 histone modifications was found to be (0.022-0.539). Specifically, H3k27ac, H3k4me1, H3k4me2, and H3k4me3 were found to be highly correlated, high levels of which have been detected surrounding transcription start sites [@barski2007high]. Likewise, for the 79 TFBS, the range of Jaccard indices was (<0.001-0.425). Here, known mediators of loop extrusion components were found to have high pairwise correlations including CTCF, RAD21, and SMC3 [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Similar profiles of Jaccard indices were exhibited across the sets of annotations for the K562 cell line (Supplementary Table 4). [??? So what? Why did you talk about it?; !!!added implication]. The high Jaccard indices among select sets of genomic annotations indicated the colocalization of annotations known to be enriched at TAD boundaries. This gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space of all annotations while retaining the possibly correlated annotations that could be predictive of TAD boundary formation.
-->

To establish the feature space for all downstream models, we constructed 3 different predictor-types. These predictor-type specific annotations represented the data matrix, $X = \{X_1, X_2, ..., X_p\}$, where $p$ was the complete set of functional annotations. Overlap counts was the least granular predictor-type considered, simply defined by the number of annotation-specific regions located within each genomic bin (Figure 2A). Overlap percents was a more refined predictor-type, defined as the total percent overlap between each annotation-specific region and genomic bins (Figure 2B). Lastly, distances, the most granular predictor-type, described the spatial relationship between the center of each genomic bin to the center of the nearest annotation-specific region (Figure 2C). A $log_{2}}$ transformation was used as a normalization technique in order to account for the skewness present in the distance densities (Supplementary Figures 1-2).

To systematically summarize and compare our results of data driven random forest models, we created an ensemble framework (Figure 3). Each system in the framework was assumed to be composed of a model constructed from a particular cell line, $L = \{GM12878, K562\}$. Within each cell line we evaluated 4 data resolutions, $R = \{10 kb, 25 kb, 50 kb, 100 kb\}$. We then considered 3 separate predictor types as the feature space including $P = \{OC, OP, Distance\}$. To assess the severity of class imbalance, 4 re-sampling techniques were implemented prior to modelling, including $B = \{None, RUS, ROS, SMOTE\}$. Each combination of the above parameters was implemented on chromosome specific sections of the linear genome. Performances were then aggregated across each of the 21 chromosomes (omitting chromosome 9). Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling 2016 ($2 \times 4 \times 3 \times 4 \times 21$) separate models. We report our main results for the GM12878 cell line and have included results from the K562 cell line in the supplementary materials.
